---
title: "Code optimization"
author: "Formation R avancé. INRA."
date: "Decembre 2016."
knit: (function(inputFile, encoding) { 
        rmarkdown::render(
          inputFile,
          encoding = encoding, 
          output_dir = "../Day2",
          output_format = "beamer_presentation"
        ) 
      })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Loops and vectorized operations

## Computing a likelihood

A likelihood often involves terms such as $\log(\prod_{i=1}^n x_i)$ for the vector of observations $\mathbf x$.

However, for numerical accuracy it is more convenient to compute $\sum_i \log(x_i)$ instead.

- Take 
    ```{r long-vector}
    x <- runif(1e5)
    ```

    and compute $\sum_i \log(x_i)$
    
## sum-log alternative functions
    
```{r sum-log-alternatives}
funs <- list(
  loop = function(x) {
    slx <- 0
    for( i in seq_along(x) ) {
      slx <- slx + log(x[i])
    }
  },
  vectorized = function(x) sum(log(x))
)


```

## sum-log benchmarks

```{r sum-log-times}
time_f <- function(f, x) system.time(f(x))
sapply(funs, time_f, x)
```


## Take-away message

- Here, `log()` is a *vectorized function* (i.e. `log(vec) = vec(logs)`)

Rather than looping, \alert{exploit vectorized operations} as much as you can

(see Programming - [Vectorized Operations](../support/Programming.html#vectorized-operations))


## Growing computations

- Compute the squares of the first $N$ integers


## Growing alternative functions

```{r growing-alternative-functions}
funs <- list(
  grow_vec = function(n) {
    a <- NULL
    for(i in 1:n) a <- c(a, i^2)
    a},
  preal_vec = function(n) {
    a <- numeric(n)
    for(i in 1:n) a[i] <- i^2
    a},
  vectorized = function(n) {
    a <- (1:n)^2
    a}
)

```


## Growing benchmarks

```{r growing-benchmarks}
sapply(funs, time_f, x = 1e4)
```


## Take-away messages

- Use \alert{vectorized} operations when possible (did I say that already?)

- If you need/want a loop, \alert{preallocate} space, and minimize type
conversion and function calls within the loop



## Vectorizing with `*apply()`

The `apply()` family of functions allows building a vectorized version of a
non-vectorized functions

- Given a matrix `x` such as

```{r alternative-arithmetic-means}
N <- 1e2
x <- t( 1:N + matrix(rnorm(N^2), N) )
```

1. Can you understand what does the code do?

2. Write a function to compute the row sums


## Alternative functions for computing row-sums

```{r alternative-functions-colmeans, warning=FALSE}
funs <- list(
  loop = function(x) {
    # pre-allocate space and type
    cm <- vector("double", length = nrow(x))
    for (i in seq_len(nrow(x))) {
      cm[i] <- mean(x[i, ])
    }
    return(cm)
  },
  apply = function(x) apply(x, 1, mean),
  vectorized = function(x) rowMeans(x)
)

```


## Comparison of performance with `system.time()`

```{r performance-system-time}
sapply(funs, time_f, x)  # too fast
```

- All alternatives are **too fast** to notice differences

- Even if noticeable, results would be **noisy**

- Idea: **repeat** each approach many times, and compare distributions


## Comparison of performance with the `microbenchmark` package

```{r performance-microbenchmark, fig.width = 4, fig.height = 2}
library(microbenchmark); library(ggplot2)
mb <- with(funs,
           microbenchmark(loop(x), apply(x), vectorized(x)))
autoplot(mb) + expand_limits(y=100)
```



## Take-away message

- `*apply()` is not **necessarily** faster than `for` loops (althoug it often
is)

- Use native vectorized functions when possible (`rowSums()`, `colSums()`,
`rowMeans()`, `colMeans()`)


## Think vector

> Vectorize your code

- Avoid `for` loops (mainly for expressiveness)

    - but if you use, then preallocate space

- Master the `*apply()` family and/or the `plyr` package

- Yet, prefer the most specialized (vectorized) function available (e.g. `row/colSums()`, `row/colMeans()`)

- There are some specific situations where loops are more natural/convenient 
    (See [Advanced R. \S 11.6](http://adv-r.had.co.nz/Functionals.html\#functionals-not))


## Don't overoptimize

- Unless you know in advance that performance will be a problem, focus on
writing robust, reliable, clear and easy to maintain code

- Optimize only \alert{if needed}, \alert{where needed}


# Profiling


## Find out *where* optimization is needed

```{r profiling-example, eval = FALSE}
library(profvis)
profvis({
  data(diamonds, package = "ggplot2")
  
  plot(price ~ carat, data = diamonds)
  m <- lm(price ~ carat, data = diamonds)
  abline(m, col = "red")
})
```


## Further reading

(see Programming - [Profiling](../support/Programming.html#profiling))



# Compiled code


- `compiler::cmpfun()` byte-compile R functions

- `Rcpp::cppFunction()` write and compile C++ code from R on the fly



## Compile example

\small

```{r compilers-example}
sumR <- function(x) {
  total <- 0
  for (i in seq_along(x)) {
    total <- total + x[i]
  }
  total}

sum_bytec <- compiler::cmpfun(sumR)

sumC <- Rcpp::cppFunction('double sumC(NumericVector x) {
    int n = x.size(); double total = 0;
    for(int i = 0; i < n; ++i) {
      total += x[i];
    }
    return total;  }')

sum_vec <- function(x) sum(x)
```

## Compilers benchmark

```{r compilers-benchmark, fig.width = 4, fig.height = 2}
library(microbenchmark); library(ggplot2)
x <- runif(1e3)
autoplot(
    microbenchmark(sumR(x), sum_bytec(x), sumC(x), sum_vec(x))
  )
```

## Further reading

- see Programming
    - [Byte code compilation](../support/Programming.html#byte-code-compilation-compilercmpfun)

    - [Interfacing with C(++)](../support/Programming.html#interfacing-with-c-package-rcpp)

